---
title: "Coding_ranges"
author: "Grace Carey"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(tidyverse)
library(dplyr)
library(tidyr)
```

First read in your annotated reference genome and the variants you are interested in. 
Make sure your files are quality filtered and ready to go. 

```{r, get files ready and quality filter}

annotat <- read.csv("MG1655_flye_annotation.csv") #this file contains the annotation for the MG1655 reference strain assembled/annotated genome
variants_df1 <- read.csv("combined_variations.csv")

# head(annotat)

variants_df <- dplyr::filter(variants_df1, Score >= 20)

cds_data <- annotat[annotat$type == "CDS", ]
names(cds_data)[names(cds_data) == "function."] <- "region_function"

```

Then use GenomicRanges to find where coding ranges (CDS) are found in the reference genome. 

```{r, find CDS ranges}
# Ensure start is always less than or equal to stop
cds_data$start_fixed <- pmin(cds_data$start, cds_data$stop)
cds_data$stop_fixed <- pmax(cds_data$start, cds_data$stop)


library(GenomicRanges)

cds_gr <- GRanges(
  seqnames = cds_data$contig_id,
  ranges = IRanges(start = cds_data$start_fixed, end = cds_data$stop_fixed),
  strand = cds_data$strand
)

mcols(cds_gr)$feature_id <- cds_data$feature_id
mcols(cds_gr)$region_function <- cds_data$region_function

```

Next find which variants of interest exist within a coding region (CDS). 

```{r, find which variants are in cds}

library(GenomicRanges)

variants_gr <- GRanges(
  seqnames = variants_df$Contig,
  ranges = IRanges(start = variants_df$Pos, width = 1)
)
# unique(seqnames(variants_gr)) #use this to check what the contig name is for the variants data

# Rename contig name in cds_gr to match the variant contig name
seqlevels(cds_gr) <- "562.163959.con.0001"

hits <- findOverlaps(variants_gr, cds_gr) #which variants are in a CDS?

# Extract overlapping variants from variants_df
variants_in_cds <- variants_df[queryHits(hits), ] #get the variants in CDS

# Extract CDS metadata (e.g., feature_id, region_function)
cds_hits <- cds_gr[subjectHits(hits)]
cds_meta <- as.data.frame(mcols(cds_hits))

# Combine both
variants_in_cds$feature_id <- cds_meta$feature_id
variants_in_cds$region_function <- cds_meta$region_function #now you have the desired variants


# write.csv(variants_in_cds, "variants_in_CDS.csv", row.names = FALSE) #can write to csv if needed. 

```

And finally find which variants exist within 500 basepairs (upstream or downstream) of a coding region. 

```{r, variants within 500bp of CDS}

# Expand CDS ranges by 500bp on either side
cds_gr_expanded <- resize(cds_gr, width = width(cds_gr) + 1000, fix = "center")

# Find overlaps between variants and the expanded CDS ranges
hits_expanded <- findOverlaps(variants_gr, cds_gr_expanded)

# Extract overlapping variants
variants_near_cds <- variants_df[queryHits(hits_expanded), ] # get the variants near CDS

# Extract expanded CDS metadata
cds_hits_expanded <- cds_gr_expanded[subjectHits(hits_expanded)]
cds_meta_expanded <- as.data.frame(mcols(cds_hits_expanded))

# Combine both
variants_near_cds$feature_id <- cds_meta_expanded$feature_id
variants_near_cds$region_function <- cds_meta_expanded$region_function # variants near CDS regions

# Optionally, save to a CSV
# write.csv(variants_near_cds, "variants_near_CDS.csv", row.names = FALSE)

# Split Var column by comma and expand the rows accordingly
variants_df_split <- variants_near_cds %>%
  separate_rows(Var, sep = ",")  # This will separate Var into individual rows

# Check the result
print(variants_df_split)
```

Now that we have the variants within 500bp of a coding region, I will see which (non-synonymous) ones are unique to AZ cultures, which are unique to control cultures, and which are common to both. Unique will be defined as:
Unique in position AND sequence. Since multiple variants can occur at a single position. 


```{r}
library(dplyr)
library(writexl)

# Read the key file
Key <- read.csv("SNPs_key.csv")

# Merge with variants data
merged_variants <- merge(variants_df_split, Key, by = "Library_name", all.x = TRUE)

# Create Pos_var to uniquely identify variants
merged_variants <- merged_variants %>%
  mutate(Pos_var = paste(Pos, Var, sep = "_"))

# Separate into AZ and Control groups
azithro <- merged_variants %>%
  filter(Evolution_conditions == "Azithromycin")

noazithro <- merged_variants %>%
  filter(Evolution_conditions == "NoAzithromycin")

# Identify unique and common Pos_vars
az_posvars <- azithro$Pos_var
noaz_posvars <- noazithro$Pos_var

common_posvars <- intersect(az_posvars, noaz_posvars)
az_unique_posvars <- setdiff(az_posvars, noaz_posvars)
control_unique_posvars <- setdiff(noaz_posvars, az_posvars)

# Filter each category to non-synonymous
variants_unique_toAZ_filtered <- azithro %>%
  filter(Pos_var %in% az_unique_posvars, Type != "Synon")

variants_unique_toControl_filtered <- noazithro %>%
  filter(Pos_var %in% control_unique_posvars, Type != "Synon")

variants_common_filtered <- merged_variants %>%
  filter(Pos_var %in% common_posvars, Type != "Synon")
```



```{r}

# Function to create Excel sheet list
make_sheet_list <- function(df) {
  unique_gene_ids <- df %>%
    select(Pos, Library_name, Gene_ID, Function, region_function, feature_id) %>%
    distinct()

  list(
    All_Data = unique_gene_ids,
    Unique_Gene_ID = unique(select(unique_gene_ids, Gene_ID)),
    Unique_Function = unique(select(unique_gene_ids, Function)),
    Unique_Region_Function = unique(select(unique_gene_ids, region_function)),
    Unique_Feature_ID = unique(select(unique_gene_ids, feature_id))
  )
}
# 
# # Create and write Excel files
# write_xlsx(make_sheet_list(variants_unique_toAZ_filtered), 
#            path = "unique_gene_info_AZ.xlsx")
# 
# write_xlsx(make_sheet_list(variants_unique_toControl_filtered), 
#            path = "unique_gene_info_Control.xlsx")
# 
# write_xlsx(make_sheet_list(variants_common_filtered), 
#            path = "common_gene_info.xlsx")


```

```{r}
# variantsredundant <- read.csv("All_variants.csv")
# 
# var_distinct <- distinct(variantsredundant)

# write.csv(var_distinct, "unique_featureids_filtered_all.csv")

```
