---
title: "Coding_ranges"
author: "Grace Carey"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(tidyverse)
library(dplyr)
library(tidyr)
library(readxl)
```

# Important note

This R script and the output files are specifically examining variants present in S07 (final sampling) samples. For a complete

First read in your annotated reference genome and the variants you are interested in. 
Make sure your files are quality filtered and ready to go. 


```{r, get files ready and quality filter}

annotat <- read.csv("MG1655_flye_annotation.csv") #this file contains the annotation for the MG1655 reference strain assembled/annotated genome
variants_df1 <- read.csv("combined_variations.csv")

# head(annotat)

variants_df <- dplyr::filter(variants_df1, Score >= 20)

cds_data <- annotat[annotat$type == "CDS", ]
names(cds_data)[names(cds_data) == "function."] <- "region_function"

```

Then use GenomicRanges to find where coding ranges (CDS) are found in the reference genome. 

```{r, find CDS ranges}
# Ensure start is always less than or equal to stop
cds_data$start_fixed <- pmin(cds_data$start, cds_data$stop)
cds_data$stop_fixed <- pmax(cds_data$start, cds_data$stop)


library(GenomicRanges)

cds_gr <- GRanges(
  seqnames = cds_data$contig_id,
  ranges = IRanges(start = cds_data$start_fixed, end = cds_data$stop_fixed),
  strand = cds_data$strand
)

mcols(cds_gr)$feature_id <- cds_data$feature_id
mcols(cds_gr)$region_function <- cds_data$region_function

```

Next find which variants of interest exist within a coding region (CDS). 

```{r, find which variants are in cds}

library(GenomicRanges)

variants_gr <- GRanges(
  seqnames = variants_df$Contig,
  ranges = IRanges(start = variants_df$Pos, width = 1)
)
# unique(seqnames(variants_gr)) #use this to check what the contig name is for the variants data

# Rename contig name in cds_gr to match the variant contig name
seqlevels(cds_gr) <- "562.163959.con.0001"

hits <- findOverlaps(variants_gr, cds_gr) #which variants are in a CDS?

# Extract overlapping variants from variants_df
variants_in_cds <- variants_df[queryHits(hits), ] #get the variants in CDS

# Extract CDS metadata (e.g., feature_id, region_function)
cds_hits <- cds_gr[subjectHits(hits)]
cds_meta <- as.data.frame(mcols(cds_hits))

# Combine both
variants_in_cds$feature_id <- cds_meta$feature_id
variants_in_cds$region_function <- cds_meta$region_function #now you have the desired variants


# write.csv(variants_in_cds, "variants_in_CDS.csv", row.names = FALSE) #can write to csv if needed. 

```

And finally find which variants exist within 500 basepairs (upstream or downstream) of a coding region. 

```{r, variants within 500bp of CDS}

# Expand CDS ranges by 500bp on either side
cds_gr_expanded <- resize(cds_gr, width = width(cds_gr) + 1000, fix = "center")

# Find overlaps between variants and the expanded CDS ranges
hits_expanded <- findOverlaps(variants_gr, cds_gr_expanded)

# Extract overlapping variants
variants_near_cds <- variants_df[queryHits(hits_expanded), ] # get the variants near CDS

# Extract expanded CDS metadata
cds_hits_expanded <- cds_gr_expanded[subjectHits(hits_expanded)]
cds_meta_expanded <- as.data.frame(mcols(cds_hits_expanded))

# Combine both
variants_near_cds$feature_id <- cds_meta_expanded$feature_id
variants_near_cds$region_function <- cds_meta_expanded$region_function # variants near CDS regions

# Optionally, save to a CSV
# write.csv(variants_near_cds, "variants_near_CDS.csv", row.names = FALSE)

# Split Var column by comma and expand the rows accordingly
variants_df_split <- variants_near_cds %>%
  separate_rows(Var, sep = ",")  # This will separate Var into individual rows

# Check the result
print(variants_df_split)
```

Now that we have the variants within 500bp of a coding region, I will see which (non-synonymous) ones are unique to AZ cultures, which are unique to control cultures, and which are common to both. Unique will be defined as:
Unique in position AND sequence. Since multiple variants can occur at a single position. 


```{r}
library(dplyr)
library(writexl)

# Read the key file
Key <- read.csv("SNPs_key.csv")

# Merge with variants data
merged_variants <- merge(variants_df_split, Key, by = "Library_name", all.x = TRUE)

# Create Pos_var to uniquely identify variants
merged_variants <- merged_variants %>%
  mutate(Pos_var = paste(Pos, Var, sep = "_"))

merged_variantsnonsyn <- filter(merged_variants, Type != "Synon")


```

```{r}

#I am commenting this out so that it doesn't keep adding pipes to the feature_id column. This code does work, just uncomment when ready to use. 
# blastresults <- read.csv("filteredblastinfo52125.csv") #these blast results are the result of blasting the unique variants against the nucleotide database to try and find more about what they actually are. After blasting they were quality filtered. These are the quality filtered blast results. 
# 
# merged_variantsnonsyn$feature_id <- paste0(merged_variantsnonsyn$feature_id, "|")
# 
# blastedvariants <- merge(blastresults, merged_variantsnonsyn,
#                          by.x = "qseqid", by.y = "feature_id", all = TRUE)

# write.csv(blastedvariants, "variantswithblastfiltered.csv")

```


```{r}

# blastedvariants1 <- read.csv("variantswithblastfiltered.csv")
# blastedvariants <- filter(blastedvariants1, Timepoint == "S07")
blastedvariants <- read.csv("variantswithblastfiltered.csv")


#in this dataframe, my own personal variant ids are in qseqid, and the feature IDs that BV BRC identified are in result.feature.id
blastedvariants <- blastedvariants %>%
  mutate(Gene.Symbol = ifelse(Gene.Symbol == "" | is.na(Gene.Symbol), result.feature.id, Gene.Symbol))

length(unique(blastedvariants$feature_id))
length(unique(blastedvariants$qseqid))
sum(is.na(blastedvariants$subclass))
blastedvariants$Gene.Symbol <- as.character(blastedvariants$Gene.Symbol)
blastedvariants$Gene.Symbol[blastedvariants$Gene.Symbol == "fig|1133852.3.peg.3993"] <- "acrA" #checked the fig id in BVBRC and blasted nucleotides to check gene ID
blastedvariants$Gene.Symbol[blastedvariants$Gene.Symbol == "fig|386585.9.peg.1523"] <- "ymdA" #checked the fig id in BVBRC and blasted nucleotides to check gene ID


```



```{r}
custom_order <- c("AZ01_S01", "AZ01_S04", "AZ01_S07",
                  "AZ02_S01", "AZ02_S04", "AZ02_S07",
                  "AZ03_S01", "AZ03_S04", "AZ03_S07",
                  "AZ04_S01", "AZ04_S04", "AZ04_S07",
                  "AZ05_S01", "AZ05_S04", "AZ05_S07",
                  "CN01_S01", "CN01_S04", "CN01_S07",
                  "CN02_S01", "CN02_S04", "CN02_S07")



blastedvariants$Sample <- factor(blastedvariants$Sample, levels = custom_order)
ggplot(blastedvariants, aes(x = Sample, fill = Type)) +
  geom_bar(position = "stack") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Count of Variant Types per Sample",
    x = "Sample",
    y = "Count",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) 
    # geom_vline(xintercept = c(0.5,3.5,6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid")

# ggsave("variant_counts_bars.png", width = 8, height = 6, dpi = 300)


```





```{r}


# Define colors for subclasses - you can customize these hex codes
subclass_colors <- c(
  "Capsule and Slime layer" = "#1f77b4",
  "Multidrug efflux systems" = "#ff7f0e",
  "Branched-chain amino acids" = "#2ca02c",
  "Fermentation" = "#d62728",
  "Transcription" = "#9467bd",
  "Ribosome biogenesis" = "#8c564b",
  "RNA processing and modification" = "#e377c2",
  "Protein folding" = "#7f7f7f",
  "Aromatic amino acids and derivatives" = "#bcbd22",
  "No_subclass" = "gray80"  # Add this line for 'None'
)
```

```{r}
blastedvariants <- filter(blastedvariants, Start != "NA")

blastedvariants <- blastedvariants %>%
  mutate(
    Gene_with_sstart = paste0(Gene.Symbol, " ", format(Start, scientific = TRUE) )
  ) %>%
  arrange(Start) %>%
  mutate(Gene_with_sstart = factor(Gene_with_sstart, levels = unique(Gene_with_sstart)))



```

```{r}
library(dplyr)

# Create combined identifier
blastedvariants2 <- blastedvariants %>%
  dplyr::mutate(
    Gene.Symbol = as.character(Gene.Symbol),
    Type = as.character(Type),
    Gene.Symbol.Type = paste(Gene.Symbol, Type, sep = "_")
  )


# Remove rows with missing treatment info, if any
blastedvariants_clean <- blastedvariants2 %>%
  filter(!is.na(Evolution_conditions))

# Map gene to treatment combinations
gene_treatment_map <- blastedvariants_clean %>%
  distinct(Gene.Symbol.Type, Evolution_conditions)

# Summarize treatments per gene
gene_type_counts <- gene_treatment_map %>%
  group_by(Gene.Symbol.Type) %>%
  summarise(
    conditions = list(sort(unique(Evolution_conditions))),
    .groups = "drop"
  )

# Classify based on condition combinations
gene_type_counts <- gene_type_counts %>%
  rowwise() %>%
  mutate(
    Uniqueness = case_when(
      setequal(conditions, "Azithromycin") & length(conditions) == 1 ~ "Unique to azithromycin",
      setequal(conditions, "NoAzithromycin") & length(conditions) == 1 ~ "Unique to control",
      setequal(conditions, c("Azithromycin", "NoAzithromycin")) ~ "Present in both",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()


# Merge Uniqueness back into full data
blastedvariants2 <- blastedvariants2 %>%
  left_join(
    gene_type_counts %>% select(Gene.Symbol.Type, Uniqueness),
    by = "Gene.Symbol.Type"
  )

blastedvariants2 <- blastedvariants2 %>%
  mutate(Gene.Symbol = recode(Gene.Symbol,
    "fig|1133852.3.peg.1097" = "Unknown 1",
    "fig|386585.9.peg.3878" = "Unknown 2"
  ))

```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Desired sample order
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01", "AZ01_S07","AZ01_S04","AZ01_S01")

# Fill in missing Gene.Symbol–Sample combinations
blastedvariants2 <- blastedvariants2 %>%
  complete(Gene.Symbol, Sample = desired_samples)

# Reapply factor levels explicitly after complete()
blastedvariants2$Sample <- factor(blastedvariants2$Sample, levels = desired_samples)


variantgenetypeall <- ggplot(blastedvariants2, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "gray50", size = 0.5, na.rm = FALSE) +
scale_fill_manual(
  values = c(
    "Insertion" = "#71a80d",
    "Nonsyn" = "#4caaea",
    "Deletion" = "#d75b43"
  ),
  na.value = "white"  # <--- this sets the color for NA fills
) +
  labs(
    title = "Genes with variants",
    x = "Gene",
    y = "Sample",
    fill = "Type"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10)
  )



variantgenetypeall


# ggsave(
#   filename = "variantgenetypeall.png",     # or "micwithline.pdf", "micwithline.svg"
#   plot = variantgenetypeall,
#   width = 8,       # width in inches
#   height = 6,      # height in inches
#   dpi = 300        # resolution
# )


```

```{r}
# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # Desired sample order
# # desired_samples <- c(
# #   "CN02_S07", "CN01_S07",
# #   "AZ05_S07", "AZ04_S07",
# #   "AZ03_S07", "AZ02_S07", "AZ01_S07"
# # )
# 
# # # Fill in missing Gene.Symbol–Sample combinations
# # blastedvariants2 <- blastedvariants2 %>%
# #   complete(Gene.Symbol, Sample = desired_samples)
# blastedvariantshatch <- blastedvariants2
# # Reapply factor levels explicitly after complete()
# blastedvariantshatch$Sample <- factor(blastedvariantshatch$Sample, levels = desired_samples)
# 
# # Step 1: Create a gene order based on the Start position
# gene_order <- blastedvariantshatch %>%
#   group_by(Gene.Symbol) %>%
#   summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
#   arrange(Start) %>%
#   pull(Gene.Symbol)
# 
# # Step 2: Apply that order to the Gene.Symbol factor
# blastedvariantshatch <- blastedvariantshatch %>%
#   mutate(Gene.Symbol = factor(Gene.Symbol, levels = gene_order))
# 
# library(ggpattern) 
# 
# variantgenetypeallhatch <- ggplot(blastedvariantshatch, aes(x = factor(Gene.Symbol), y = Sample)) +
#   geom_tile_pattern(
#     aes(
#       fill = Type,
#       pattern = ifelse(Uniqueness == "Present in both", "crosshatch", "none")
#     ),
#     pattern_fill = "black",
#     pattern_angle = 45,
#     pattern_density = 0.2,
#     pattern_spacing = 0.02,
#     color = "gray50",
#     size = 0.5
#   ) +
#   scale_fill_manual(
#     values = c(
#       "Insertion" = "#71a80d",
#       "Nonsyn" = "#4caaea",
#       "Deletion" = "#d75b43"
#     ),
#     na.value = "white"
#   ) +
#   scale_pattern_manual(values = c("none" = NA, "crosshatch" = "crosshatch")) +
#   labs(
#     title = "Genes with variants unique to AZ or control",
#     x = "Gene",
#     y = "Sample",
#     fill = "Type",
#     pattern = "Uniqueness"
#   ) +
#   theme_minimal() +
#   guides(pattern="none") +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 8))
# 
# variantgenetypeallhatch
# 
# 
# # ggsave(
# #   filename = "variantgenetypeallhatch.png",     # or "micwithline.pdf", "micwithline.svg"
# #   plot = variantgenetypeallhatch,
# #   width = 9,       # width in inches
# #   height = 6,      # height in inches
# #   dpi = 300        # resolution
# # )


```
```{r, fig.height = 20, fig.width=10}
# 
# library(ggplot2)
# library(dplyr)
# library(ggpattern)
# library(ggh4x)
# 
# # Factor level setup (Sample and Gene.Symbol)
# blastedvariantshatch <- blastedvariants2
# blastedvariantshatch <- filter(blastedvariantshatch, Timepoint != "NA")
# 
# desired_samples <- c(
#   "AZ01_S01", "AZ01_S04", "AZ01_S07",
#   "AZ02_S01", "AZ02_S04", "AZ02_S07",
#   "AZ03_S01", "AZ03_S04", "AZ03_S07",
#   "AZ04_S01", "AZ04_S04", "AZ04_S07",
#   "AZ05_S01", "AZ05_S04", "AZ05_S07",
#   "CN01_S01", "CN01_S04", "CN01_S07",
#   "CN02_S01", "CN02_S04", "CN02_S07"
# )
# 
# blastedvariantshatch <- blastedvariantshatch %>%
#   mutate(
#     Sample = factor(Sample, levels = desired_samples),
#     Culture = factor(Culture),
#     Timepoint = factor(Timepoint)
#   )
# 
# gene_order <- blastedvariantshatch %>%
#   group_by(Gene.Symbol) %>%
#   summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
#   arrange(Start) %>%
#   pull(Gene.Symbol)
# 
# blastedvariantshatch <- blastedvariantshatch %>%
#   mutate(Gene.Symbol = factor(Gene.Symbol, levels = gene_order))
# 
# variantgenetypeallhatch <- ggplot(blastedvariantshatch, aes(x = Gene.Symbol, y = Timepoint)) +
#   geom_tile_pattern(
#     aes(
#       fill = Type,
#       pattern = ifelse(Uniqueness == "Present in both", "crosshatch", "none")
#     ),
#     pattern_fill = "black",
#     pattern_angle = 45,
#     pattern_density = 0.2,
#     pattern_spacing = 0.02,
#     color = "gray50",
#     size = 0.5
#   ) +
#   facet_grid(Culture ~ ., 
#              labeller = label_value,
#              scales = "fixed",
#              space = "fixed") +
#   scale_fill_manual(
#     values = c(
#       "Insertion" = "#71a80d",
#       "Nonsyn" = "#4caaea",
#       "Deletion" = "#d75b43"
#     ),
#     na.value = "white"
#   ) +
#   scale_pattern_manual(values = c("none" = NA, "crosshatch" = "crosshatch")) +
#   guides(pattern = "none") +
#   labs(
#     title = "Genes with variants unique to AZ or control",
#     x = "Gene",
#     y = "Timepoint",
#     fill = "Type",
#     pattern = "Uniqueness"
#   ) +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(angle = 0, size = 8),  # horizontal timepoint labels
#     strip.text.y = element_text(size = 10, angle = 0),
#     strip.background = element_rect(fill = "gray90", color = "lightgray"),
#     strip.placement = "outside",
#     panel.spacing = unit(0.5, "lines")
#   )
# 
# variantgenetypeallhatch
```

```{r, fig.height=20, fig.width=10}


# ggplot(blastedvariantshatch, aes(x = Gene.Symbol, y = Timepoint)) +
#   geom_tile_pattern(
#     aes(
#       fill = Type,
#       pattern = ifelse(Uniqueness == "Present in both", "crosshatch", "none")
#     ),
#     pattern_fill = "black",
#     pattern_angle = 45,
#     pattern_density = 0.2,
#     pattern_spacing = 0.02,
#     color = "gray50",
#     size = 0.5
#   ) +
#   facet_grid(Culture ~ ., switch = "y", labeller = label_value) +
#   scale_y_discrete(position = "right") +
#   scale_fill_manual(values = c("Insertion" = "#71a80d", "Nonsyn" = "#4caaea", "Deletion" = "#d75b43")) +
#   scale_pattern_manual(values = c("none" = NA, "crosshatch" = "crosshatch")) +
#   guides(pattern = "none") +
#   labs(
#     x = "Gene",
#     y = "Timepoint",
#     fill = "Type",
#     pattern = "Uniqueness"
#   ) +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, size = 7, vjust = 0.5),
#     axis.title.y.right = element_text(angle = 270, vjust = 0.5),
#     axis.title.y = element_blank(),
#     axis.text.y.right = element_text(size = 8),
#     axis.text.y = element_blank(),
#     axis.ticks.y = element_blank(),
# 
#     strip.placement = "outside",
#     strip.text.y.right = element_text(angle = 0, size = 10),
#     panel.spacing = unit(0.5, "lines")
#   )


```

```{r, fig.width=10, fig.height=20}
ggplot(blastedvariantshatch, aes(x = Gene.Symbol, y = Sample, fill = Type)) +  # y = Sample needed here
  geom_tile_pattern(
    aes(
      pattern = ifelse(Uniqueness == "Present in both", "crosshatch", "none")
    ),
    pattern_fill = "black",
    pattern_angle = 45,
    pattern_density = 0.2,
    pattern_spacing = 0.02,
    color = "gray50",
    size = 0.5,
    height = 0.9
  ) +
  facet_nested(Culture + Timepoint ~ .,
               labeller = label_value,
               scales = "fixed",
               space = "fixed") +
  scale_fill_manual(
    values = c(
      "Insertion" = "#71a80d",
      "Nonsyn" = "#4caaea",
      "Deletion" = "#d75b43"
    ),
    na.value = "white"
  ) +
  scale_pattern_manual(values = c("none" = NA, "crosshatch" = "crosshatch")) +
  guides(pattern = "none") +
  labs(
    title = "Genes with variants unique to AZ or control",
    x = "Gene",
    y = NULL,  # hide y-axis label
    fill = "Type",
    pattern = "Uniqueness"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_blank(),    # hide y axis text
    axis.ticks.y = element_blank(),   # hide y axis ticks
    strip.background = element_rect(fill = "gray90", color = "lightgray"),
    strip.placement = "outside",
    panel.spacing = unit(0.5, "lines")
  )

```

```{r}
# Desired sample order
desired_samples <- c(
  "CN02_S07", "CN01_S07",
  "AZ05_S07", "AZ04_S07",
  "AZ03_S07", "AZ02_S07", "AZ01_S07"
)

blastedvariantsfiltered <- filter(blastedvariants2, Timepoint == "S07")

# Fill in missing Gene.Symbol–Sample combinations
blastedvariantsfiltered <- filter(blastedvariantsfiltered, Uniqueness != "Present in both")

blastedvariantsfiltered <- blastedvariantsfiltered %>%
  complete(Gene.Symbol, Sample = desired_samples)
blastedvariantsfiltered <- filter(blastedvariants2, Timepoint == "S07")
# Reapply factor levels explicitly after complete()
blastedvariantsfiltered$Sample <- factor(blastedvariantsfiltered$Sample, levels = desired_samples)
library(forcats)

# blastedvariantsfiltered <- blastedvariantsfiltered %>%
#   mutate(Gene.Symbol = fct_infreq(Gene.Symbol))
# Step 1: Create a gene order based on the Start position
gene_order <- blastedvariantsfiltered %>%
  group_by(Gene.Symbol) %>%
  summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
  arrange(Start) %>%
  pull(Gene.Symbol)

# Step 2: Apply that order to the Gene.Symbol factor
blastedvariantsfiltered <- blastedvariantsfiltered %>%
  mutate(Gene.Symbol = factor(Gene.Symbol, levels = gene_order))


variantgenetypeall2 <- ggplot(blastedvariantsfiltered, aes(x = factor(Gene.Symbol), y = Sample)) +
  geom_tile(aes(fill = Type), color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c(
      "Insertion" = "#71a80d",
      "Nonsyn" = "#4caaea",
      "Deletion" = "#d75b43"
    ),
    na.value = "white"
  ) +
  labs(
    title = "Genes with variants unique to AZ or control",
    x = "Gene",
    y = "Sample",
    fill = "Type"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust= 0.5, size = 7),
    axis.text.y = element_text(size = 10)
  )


variantgenetypeall2


# ggsave(
#   filename = "variantgenetypeall2.png",     # or "micwithline.pdf", "micwithline.svg"
#   plot = variantgenetypeall2,
#   width = 8,       # width in inches
#   height = 6,      # height in inches
#   dpi = 300        # resolution
# )


```

```{r, fig.height= 15, fig.width= 10}

library(ggplot2)
library(dplyr)
# Remove "NA" as character (not actual NA)
blastedvariants2na <- filter(blastedvariants2, Timepoint != "NA")
desired_samples <- c(
"AZ01_S01", "AZ01_S04", "AZ01_S07",
"AZ02_S01", "AZ02_S04", "AZ02_S07",
"AZ03_S01", "AZ03_S04", "AZ03_S07",
"AZ04_S01", "AZ04_S04", "AZ04_S07",
"AZ05_S01", "AZ05_S04", "AZ05_S07",
"CN01_S01", "CN01_S04", "CN01_S07",
"CN02_S01", "CN02_S04", "CN02_S07")

blastedvariants2na <- blastedvariants2na %>%
  mutate(Sample = factor(Sample, levels = desired_samples))

# Step 1: Create a gene order based on the Start position
gene_order <- blastedvariants2na %>%
  group_by(Gene.Symbol) %>%
  summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
  arrange(Start) %>%
  pull(Gene.Symbol)

# Step 2: Apply that order to the Gene.Symbol factor
blastedvariants2na <- blastedvariants2na %>%
  mutate(Gene.Symbol = factor(Gene.Symbol, levels = gene_order))

# Plot
ggplot(blastedvariants2na, aes(x = Gene.Symbol, y = Var_frac, fill = Type)) +
  geom_col(position = "dodge") +
  facet_grid(Sample ~ ., scales = "fixed", space = "fixed") +
  labs(
    x = "Gene",
    y = "Variant Fraction",
    fill = "Variant Type",
    title = "Variant Fraction by Gene, Type, Culture, and Timepoint"
  ) +
  theme_minimal() +
  geom_hline(yintercept = c(1,0), linewidth = 0.25) +
  theme(
  axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
  strip.text.y = element_text(angle = 0, size = 10)  # horizontal text for row strips
)


```

```{r, fig.height= 15, fig.width= 15}

library(ggplot2)
library(ggh4x)
library(dplyr)

# Filter and prepare data
blastedvariants2na <- blastedvariants2 %>%
  filter(Timepoint != "NA")

desired_samples <- c(
  "AZ01_S01", "AZ01_S04", "AZ01_S07",
  "AZ02_S01", "AZ02_S04", "AZ02_S07",
  "AZ03_S01", "AZ03_S04", "AZ03_S07",
  "AZ04_S01", "AZ04_S04", "AZ04_S07",
  "AZ05_S01", "AZ05_S04", "AZ05_S07",
  "CN01_S01", "CN01_S04", "CN01_S07",
  "CN02_S01", "CN02_S04", "CN02_S07"
)

# Order factor levels
gene_order <- blastedvariants2na %>%
  group_by(Gene.Symbol) %>%
  summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
  arrange(Start) %>%
  pull(Gene.Symbol)

blastedvariants2na <- blastedvariants2na %>%
  mutate(
    Sample = factor(Sample, levels = desired_samples),
    Gene.Symbol = factor(Gene.Symbol, levels = gene_order),
    Culture = factor(Culture),
    Timepoint = factor(Timepoint)
  )

# Plot with nested facets: rows = Sample, columns = Culture + Timepoint
varfracbars <- ggplot(blastedvariants2na, aes(x = Gene.Symbol, y = Var_frac, fill = Type)) +
  geom_col(position = "dodge") +
  facet_nested(
    Culture + Timepoint ~ . ,
    labeller = label_value,
    scales = "fixed",
    space = "fixed"
  ) +
  labs(
    x = "Gene",
    y = "Variant Fraction",
    fill = "Variant Type",
    title = "Variant Fraction by Gene, Type, Culture, Timepoint and Sample"
  ) +
  theme_minimal() +
  geom_hline(yintercept = c(0, 1), linewidth = 0.25) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    strip.text.y = element_text(angle = 0, size = 10),
    panel.spacing = unit(0.5, "lines"),
    strip.background = element_rect(fill = "gray90", color = "lightgray"),
    strip.placement = "outside"
  )

# ggsave(
#   filename = "var_frac_large.png",     # or "micwithline.pdf", "micwithline.svg"
#   plot = varfracbars,
#   width = 15,       # width in inches
#   height = 15,      # height in inches
#   dpi = 300        # resolution
# )
```

```{r}

library(ggplot2)
library(ggh4x)
library(dplyr)

# Filter and prepare data
blastedvariants2na <- blastedvariants2 %>%
  filter(Timepoint != "NA")

# Desired sample order
desired_samples <- c(
  "AZ01_S01", "AZ01_S04", "AZ01_S07",
  "AZ02_S01", "AZ02_S04", "AZ02_S07",
  "AZ03_S01", "AZ03_S04", "AZ03_S07",
  "AZ04_S01", "AZ04_S04", "AZ04_S07",
  "AZ05_S01", "AZ05_S04", "AZ05_S07",
  "CN01_S01", "CN01_S04", "CN01_S07",
  "CN02_S01", "CN02_S04", "CN02_S07"
)

# Factor ordering
gene_order <- blastedvariants2na %>%
  group_by(Gene.Symbol) %>%
  summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
  arrange(Start) %>%
  pull(Gene.Symbol)

blastedvariants2na <- blastedvariants2na %>%
  mutate(
    Sample = factor(Sample, levels = desired_samples),
    Gene.Symbol = factor(Gene.Symbol, levels = gene_order),
    Culture = factor(Culture),
    Timepoint = factor(Timepoint)
  )

# List of unique Cultures
cultures <- levels(blastedvariants2na$Culture)

# Loop over cultures and generate/save plots
for (cult in cultures) {
  df_cult <- blastedvariants2na %>% filter(Culture == cult)

  # Ensure all Gene.Symbol levels are preserved
  df_cult <- df_cult %>%
    mutate(Gene.Symbol = factor(Gene.Symbol, levels = gene_order))

  p <- ggplot(df_cult, aes(x = Gene.Symbol, y = Var_frac, fill = Type)) +
    geom_col(position = "dodge") +
    facet_nested(Timepoint ~ ., labeller = label_value, scales = "fixed", space = "fixed") +
    labs(
      title = paste("Variant Fraction -", cult),
      x = "Gene",
      y = "Variant Fraction",
      fill = "Variant Type"
    ) +
    geom_hline(yintercept = c(0, 1), linewidth = 0.25) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
      strip.text.y = element_text(angle = 0, size = 10),
      panel.spacing = unit(0.5, "lines"),
      strip.background = element_rect(fill = "gray90", color = "lightgray"),
      strip.placement = "outside"
    )

  print(p)

  # Optional: save to file
  # ggsave(
  #   filename = paste0("variant_frac_", cult, ".png"),
  #   plot = p,
  #   width = 12,
  #   height = 10,
  #   dpi = 300
  # )
}



```
```{r, fig.width = 10}

library(ggplot2)
library(ggh4x)
library(dplyr)
library(tidyr)

# Filter and prepare data
blastedvariants2na <- blastedvariants2 %>%
  filter(Timepoint != "NA")

# Desired sample order
desired_samples <- c(
  "AZ01_S01", "AZ01_S04", "AZ01_S07",
  "AZ02_S01", "AZ02_S04", "AZ02_S07",
  "AZ03_S01", "AZ03_S04", "AZ03_S07",
  "AZ04_S01", "AZ04_S04", "AZ04_S07",
  "AZ05_S01", "AZ05_S04", "AZ05_S07",
  "CN01_S01", "CN01_S04", "CN01_S07",
  "CN02_S01", "CN02_S04", "CN02_S07"
)

# Order genes by Start
gene_order <- blastedvariants2na %>%
  group_by(Gene.Symbol) %>%
  summarise(Start = min(Start, na.rm = TRUE), .groups = "drop") %>%
  arrange(Start) %>%
  pull(Gene.Symbol)

# Apply factor levels globally
blastedvariants2na <- blastedvariants2na %>%
  mutate(
    Sample = factor(Sample, levels = desired_samples),
    Gene.Symbol = factor(Gene.Symbol, levels = gene_order),
    Culture = factor(Culture),
    Timepoint = factor(Timepoint)
  )

# List of unique cultures
cultures <- levels(blastedvariants2na$Culture)

# Loop over cultures and plot
for (cult in cultures) {
  df_cult <- blastedvariants2na %>%
    filter(Culture == cult) %>%
    mutate(
      Gene.Symbol = factor(Gene.Symbol, levels = gene_order),
      Timepoint = factor(Timepoint, levels = levels(blastedvariants2na$Timepoint))
    ) %>%
    complete(Gene.Symbol, Timepoint, fill = list(Var_frac = 0))

  p <- ggplot(df_cult, aes(x = Gene.Symbol, y = Var_frac, fill = Type)) +
    geom_col(position = "dodge") +
    facet_nested(Timepoint ~ ., labeller = label_value, scales = "fixed", space = "fixed") +
    labs(
      title = paste("Variant Fraction -", cult),
      x = "Gene",
      y = "Variant Fraction",
      fill = "Variant Type"
    ) +
    geom_hline(yintercept = c(0, 1), linewidth = 0.25) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
            axis.text.y = element_text(size = 15),
      strip.text.y = element_text(angle = 0, size = 15),
      panel.spacing = unit(0.5, "lines"),
      strip.background = element_rect(fill = "gray90", color = "lightgray"),
      strip.placement = "outside"
    )

  # Save the plot
  ggsave(
    filename = paste0("variant_frac_", cult, ".png"),
    plot = p,
    width = 12,
    height = 10,
    dpi = 300
  )
}



```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# Desired sample order
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01", "AZ01_S07","AZ01_S04","AZ01_S01")

# Fill in missing Gene.Symbol–Sample combinations
blastedvariants2 <- blastedvariants2 %>%
  complete(Gene.Symbol, Sample = desired_samples)

# Reapply factor levels explicitly after complete()
blastedvariants2$Sample <- factor(blastedvariants2$Sample, levels = desired_samples)


variantsorted <- ggplot(blastedvariants2, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "gray50", size = 0.5, na.rm = FALSE) +
scale_fill_manual(
  values = c(
    "Insertion" = "#71a80d",
    "Nonsyn" = "#4caaea",
    "Deletion" = "#d75b43"
  ),
  na.value = "white"  # <--- this sets the color for NA fills
) +
  labs(
    title = "Genes with variants",
    x = "Gene",
    y = "Sample",
    fill = "Type"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10)
  )



variantsorted


# ggsave(
#   filename = "variantgenetypeall.png",     # or "micwithline.pdf", "micwithline.svg"
#   plot = variantgenetypeall,
#   width = 8,       # width in inches
#   height = 6,      # height in inches
#   dpi = 300        # resolution
# )


```





```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# --- Genes of interest ---
all_genes <- unique(c(
  "ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
  "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH",
  "sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL",
  "frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
  "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
  "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
  "yjtD", "thrA"
))

# --- Desired sample order ---
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# --- Filter, simplify, and mark presence ---
blastedvariants_simple <- blastedvariants2 %>%
  select(Gene.Symbol, Sample, qseqid) %>%
  filter(Gene.Symbol %in% all_genes) %>%
  distinct() %>%
  # Create Evolution_conditions based on Sample name
  mutate(
    Evolution_conditions = case_when(
      grepl("^AZ", Sample) ~ "Azithromycin",
      grepl("^CN", Sample) ~ "NoAzithromycin",
      TRUE ~ NA_character_
    ),
    Present = !is.na(qseqid)  # TRUE if gene is present
  ) %>%
  # Fill in missing Gene-Sample combinations
  complete(
    Gene.Symbol = all_genes,
    Sample = desired_samples,
    fill = list(Evolution_conditions = NA, Present = FALSE)
  ) %>%
  mutate(
    Sample = factor(Sample, levels = desired_samples),
    Gene.Symbol = factor(Gene.Symbol, levels = all_genes)
  )

# --- Plot ---
ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)), color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
    na.value = "white"
  ) +
  labs(
    title = "Genes with AZ or CN variants",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10)
  )

```
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# --- Genes of interest by category ---
biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
                   "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
                      "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
                      "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
                      "yjtD", "thrA")

all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))

# --- Desired sample order ---
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# --- Filter, simplify, and mark presence ---
blastedvariants_simple <- blastedvariants2 %>%
  select(Gene.Symbol, Sample, qseqid) %>%
  filter(Gene.Symbol %in% all_genes) %>%
  distinct() %>%
  # Assign categories
  mutate(
    Category = case_when(
      Gene.Symbol %in% biofilm_genes ~ "Biofilm",
      Gene.Symbol %in% efflux_genes ~ "Efflux/Transport",
      Gene.Symbol %in% persistent_genes ~ "Persistent shared",
      TRUE ~ NA_character_
    ),
    # Evolution condition based on Sample name
    Evolution_conditions = case_when(
      grepl("^AZ", Sample) ~ "Azithromycin",
      grepl("^CN", Sample) ~ "NoAzithromycin",
      TRUE ~ NA_character_
    ),
    Present = !is.na(qseqid)
  ) %>%
  # Fill in missing Gene-Sample combinations per category
  complete(
    Gene.Symbol = all_genes,
    Sample = desired_samples,
    fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
  ) %>%
  mutate(
    Sample = factor(Sample, levels = desired_samples),
    Gene.Symbol = factor(Gene.Symbol, levels = all_genes)
  )

# --- Plot with facets ---
ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)), color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
    na.value = "white"
  ) +
  labs(
    title = "Genes with AZ or CN variants by category",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 9, face = "bold")
  )



```

```{r}
# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # --- Genes of interest by category ---
# biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
#                    "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
# efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
# persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
#                       "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
#                       "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
#                       "yjtD", "thrA")
# 
# all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))
# 
# # --- Desired sample order ---
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )
# 
# # --- Filter, simplify, and mark presence ---
# blastedvariants_simple <- blastedvariants2 %>%
#   select(Gene.Symbol, Sample, qseqid) %>%
#   filter(Gene.Symbol %in% all_genes) %>%
#   distinct() %>%
#   # Assign categories
#   mutate(
#     Category = case_when(
#       Gene.Symbol %in% biofilm_genes ~ "Biofilm",
#       Gene.Symbol %in% efflux_genes ~ "Efflux/Transport",
#       Gene.Symbol %in% persistent_genes ~ "Persistent shared",
#       TRUE ~ NA_character_
#     ),
#     # Evolution condition based on Sample name
#     Evolution_conditions = case_when(
#       grepl("^AZ", Sample) ~ "Azithromycin",
#       grepl("^CN", Sample) ~ "NoAzithromycin",
#       TRUE ~ NA_character_
#     ),
#     Present = !is.na(qseqid)
#   ) %>%
#   # Fill in missing Gene-Sample combinations per category
#   complete(
#     Gene.Symbol = all_genes,
#     Sample = desired_samples,
#     fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
#   ) %>%
#   mutate(
#     Sample = factor(Sample, levels = desired_samples),
#     Gene.Symbol = factor(Gene.Symbol, levels = all_genes)
#   )
# 
# # --- Manually duplicate fadL into both categories ---
# fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
# fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux/Transport")
# fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared")
# 
# blastedvariants_simple <- blastedvariants_simple %>%
#   filter(Gene.Symbol != "fadL") %>%  # remove any old inconsistent fadL category
#   bind_rows(fadL_efflux, fadL_persistent)
# 
# # --- Plot with facets ---
# ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
#   geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)), 
#             color = "gray50", size = 0.5) +
#   scale_fill_manual(
#     values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
#     na.value = "white"
#   ) +
#   labs(
#     title = "Genes with AZ or CN variants by category",
#     x = "Gene",
#     y = "Sample",
#     fill = "Evolution condition"
#   ) +
#   facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 10),
#     strip.text.x = element_text(size = 9, face = "bold")
#   )

```
```{r}

# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # --- Genes of interest by category ---
# biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
#                    "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
# efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
# persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
#                       "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
#                       "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
#                       "yjtD", "thrA")
# 
# all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))
# 
# # --- Desired sample order ---
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )
# 
# # --- Filter, simplify, and mark presence ---
# blastedvariants_simple <- blastedvariants2 %>%
#   select(Gene.Symbol, Sample, qseqid) %>%
#   filter(Gene.Symbol %in% all_genes) %>%
#   distinct() %>%
#   # Assign categories
#   mutate(
#     Category = case_when(
#       Gene.Symbol %in% biofilm_genes ~ "Biofilm",
#       Gene.Symbol %in% efflux_genes ~ "Efflux/Transport",
#       Gene.Symbol %in% persistent_genes ~ "Persistent shared",
#       TRUE ~ NA_character_
#     ),
#     # Evolution condition based on Sample name
#     Evolution_conditions = case_when(
#       grepl("^AZ", Sample) ~ "Azithromycin",
#       grepl("^CN", Sample) ~ "NoAzithromycin",
#       TRUE ~ NA_character_
#     ),
#     Present = !is.na(qseqid)
#   ) %>%
#   # Fill in missing Gene-Sample combinations per category
#   complete(
#     Gene.Symbol = all_genes,
#     Sample = desired_samples,
#     fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
#   ) %>%
#   mutate(
#     Sample = factor(Sample, levels = desired_samples)
#   )
# 
# # --- Manually duplicate fadL into both categories ---
# fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
# fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux/Transport")
# fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared")
# 
# blastedvariants_simple <- blastedvariants_simple %>%
#   filter(Gene.Symbol != "fadL") %>%  # remove old inconsistent fadL category
#   bind_rows(fadL_efflux, fadL_persistent)
# 
# # --- Set gene ordering PER CATEGORY ---
# blastedvariants_simple <- blastedvariants_simple %>%
#   mutate(
#     Gene.Symbol = case_when(
#       Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
#       Category == "Efflux/Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
#       Category == "Persistent shared" ~ factor(Gene.Symbol, levels = persistent_genes),
#       TRUE ~ factor(Gene.Symbol, levels = all_genes) # fallback
#     )
#   )
# 
# # --- Plot with facets ---
# ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
#   geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)), 
#             color = "gray50", size = 0.5) +
#   scale_fill_manual(
#     values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
#     na.value = "white"
#   ) +
#   labs(
#     title = "Genes with AZ or CN variants by category",
#     x = "Gene",
#     y = "Sample",
#     fill = "Evolution condition"
#   ) +
#   facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 10),
#     strip.text.x = element_text(size = 9, face = "bold")
#   )

```
```{r}

# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # --- Genes of interest by category ---
# biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
#                    "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
# efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
# persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
#                       "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
#                       "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
#                       "yjtD", "thrA")
# 
# all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))
# 
# # --- Desired sample order ---
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )
# 
# # --- Filter, simplify, and mark presence ---
# blastedvariants_simple <- blastedvariants2 %>%
#   select(Gene.Symbol, Sample, qseqid) %>%
#   filter(Gene.Symbol %in% all_genes) %>%
#   distinct() %>%
#   # Assign categories
#   mutate(
#     Category = case_when(
#       Gene.Symbol %in% biofilm_genes ~ "Biofilm",
#       Gene.Symbol %in% efflux_genes ~ "Efflux/Transport",
#       Gene.Symbol %in% persistent_genes ~ "Persistent shared",
#       TRUE ~ NA_character_
#     ),
#     # Evolution condition based on Sample name
#     Evolution_conditions = case_when(
#       grepl("^AZ", Sample) ~ "Azithromycin",
#       grepl("^CN", Sample) ~ "NoAzithromycin",
#       TRUE ~ NA_character_
#     ),
#     Present = !is.na(qseqid)
#   ) %>%
#   # Fill in missing Gene-Sample combinations per category
#   complete(
#     Gene.Symbol = all_genes,
#     Sample = desired_samples,
#     fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
#   ) %>%
#   mutate(
#     Sample = factor(Sample, levels = desired_samples)
#   )
# 
# # --- Manually duplicate fadL into both categories ---
# fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
# fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux/Transport")
# # Append invisible zero-width space for persistent category
# fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared",
#                                         Gene.Symbol = paste0(Gene.Symbol, "\u200B"))
# 
# blastedvariants_simple <- blastedvariants_simple %>%
#   filter(Gene.Symbol != "fadL") %>%
#   bind_rows(fadL_efflux, fadL_persistent)
# 
# # --- Assign per-facet gene ordering ---
# blastedvariants_simple <- blastedvariants_simple %>%
#   mutate(
#     Gene.Symbol = case_when(
#       Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
#       Category == "Efflux/Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
#       Category == "Persistent shared" ~ factor(Gene.Symbol, 
#                                                 levels = c(persistent_genes[persistent_genes != "fadL"], "fadL\u200B")),
#       TRUE ~ factor(Gene.Symbol)
#     )
#   )
# 
# # --- Plot with facets ---
# ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
#   geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
#             color = "gray50", size = 0.5) +
#   scale_fill_manual(
#     values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
#     na.value = "white"
#   ) +
#   labs(
#     title = "Genes with AZ or CN variants by category",
#     x = "Gene",
#     y = "Sample",
#     fill = "Evolution condition"
#   ) +
#   facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 10),
#     strip.text.x = element_text(size = 9, face = "bold")
#   )

```

```{r}

# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # --- Genes of interest by category ---
# biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
#                    "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
# efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
# persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
#                       "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
#                       "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
#                       "yjtD", "thrA")
# 
# all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))
# 
# # --- Desired sample order ---
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )
# 
# # --- Filter, simplify, and mark presence ---
# blastedvariants_simple <- blastedvariants2 %>%
#   select(Gene.Symbol, Sample, qseqid) %>%
#   filter(Gene.Symbol %in% all_genes) %>%
#   distinct() %>%
#   # Assign categories
#   mutate(
#     Category = case_when(
#       Gene.Symbol %in% biofilm_genes ~ "Biofilm",
#       Gene.Symbol %in% efflux_genes ~ "Efflux or Transport",  # changed label with line break later
#       Gene.Symbol %in% persistent_genes ~ "Persistent shared",
#       TRUE ~ NA_character_
#     ),
#     # Evolution condition based on Sample name
#     Evolution_conditions = case_when(
#       grepl("^AZ", Sample) ~ "Azithromycin",
#       grepl("^CN", Sample) ~ "NoAzithromycin",
#       TRUE ~ NA_character_
#     ),
#     Present = !is.na(qseqid)
#   ) %>%
#   # Fill in missing Gene-Sample combinations per category
#   complete(
#     Gene.Symbol = all_genes,
#     Sample = desired_samples,
#     fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
#   ) %>%
#   mutate(
#     Sample = factor(Sample, levels = desired_samples)
#   )
# 
# # --- Manually duplicate fadL into both categories ---
# fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
# fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux or Transport")
# fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared",
#                                         Gene.Symbol = paste0(Gene.Symbol, "\u200B"))
# 
# blastedvariants_simple <- blastedvariants_simple %>%
#   filter(Gene.Symbol != "fadL") %>%
#   bind_rows(fadL_efflux, fadL_persistent)
# 
# # --- Assign per-facet gene ordering ---
# blastedvariants_simple <- blastedvariants_simple %>%
#   mutate(
#     Gene.Symbol = case_when(
#       Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
#       Category == "Efflux or Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
#       Category == "Persistent shared" ~ factor(Gene.Symbol, 
#                                                levels = c(persistent_genes[persistent_genes != "fadL"], "fadL\u200B")),
#       TRUE ~ factor(Gene.Symbol)
#     ),
#     # Add line break to facet label
#     Category = case_when(
#       Category == "Efflux or Transport" ~ "Efflux or\nTransport",
#       TRUE ~ Category
#     )
#   )
# 
# # --- Plot with facets ---
# ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
#   geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
#             color = "gray50", size = 0.5) +
#   scale_fill_manual(
#     values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
#     na.value = "white"
#   ) +
#   labs(
#     title = "Genes with AZ or CN variants by category",
#     x = "Gene",
#     y = "Sample",
#     fill = "Evolution condition"
#   ) +
#   facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 10),
#     strip.text.x = element_text(size = 9, face = "bold")
#   )
# 
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Genes of interest by category ---
biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
                   "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
                      "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
                      "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
                      "yjtD", "thrA")

all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))

# --- Desired sample order ---
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# --- Filter, simplify, and mark presence ---
blastedvariants_simple <- blastedvariants2 %>%
  select(Gene.Symbol, Sample, qseqid) %>%
  filter(Gene.Symbol %in% all_genes) %>%
  distinct() %>%
  mutate(
    Category = case_when(
      Gene.Symbol %in% biofilm_genes ~ "Biofilm",
      Gene.Symbol %in% efflux_genes ~ "Efflux or Transport",
      Gene.Symbol %in% persistent_genes ~ "Persistent shared",
      TRUE ~ NA_character_
    ),
    Evolution_conditions = case_when(
      grepl("^AZ", Sample) ~ "Azithromycin",
      grepl("^CN", Sample) ~ "NoAzithromycin",
      TRUE ~ NA_character_
    ),
    Present = !is.na(qseqid)
  ) %>%
  complete(
    Gene.Symbol = all_genes,
    Sample = desired_samples,
    fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
  ) %>%
  mutate(Sample = factor(Sample, levels = desired_samples))

# --- Duplicate fadL into both categories ---
fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux or Transport")
fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared",
                                        Gene.Symbol = paste0(Gene.Symbol, "\u200B"))

blastedvariants_simple <- blastedvariants_simple %>%
  filter(Gene.Symbol != "fadL") %>%
  bind_rows(fadL_efflux, fadL_persistent)

# --- Ordering and facet label adjustment ---
blastedvariants_simple <- blastedvariants_simple %>%
  mutate(
    Gene.Symbol = case_when(
      Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
      Category == "Efflux or Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
      Category == "Persistent shared" ~ factor(Gene.Symbol,
                                               levels = c(persistent_genes[persistent_genes != "fadL"], "fadL\u200B")),
      TRUE ~ factor(Gene.Symbol)
    ),
    Category = case_when(
      Category == "Efflux or Transport" ~ "Efflux or\nTransport",
      TRUE ~ Category
    )
  )

# --- Plot with dark facet borders ---
ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
            color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "NoAzithromycin" = "#1E88E5"),
    na.value = "white"
  ) +
  labs(
    title = "Genes with AZ or CN variants by category",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),  # dark border
    panel.border = element_rect(color = "black", fill = NA, size = 1), # thick black facet border
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 9, face = "bold"),
    strip.background = element_rect(fill = "white", color = "black", size = 1) # border around facet labels
  )
# --- Save the plot as a high-resolution PNG ---
# ggsave(
#   filename = "genes_variant_heatmap.png",   # file name
#   plot = last_plot(),                       # saves the last plot
#   width = 12,                               # width in inches
#   height = 6,                               # height in inches
#   dpi = 300                                 # resolution in dots per inch
# )
```
```{r}
# 
# library(dplyr)
# library(tidyr)
# library(ggplot2)
# 
# # --- Genes of interest by category ---
# biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
#                    "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
# efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
# persistent_genes <- c("frmA", "frmR", "yaiO", "stfP", "tfaP", "tfaE",
#                       "stfE", "pinE", "yfcA", "mepA", "fadL", "yfdF",
#                       "yifK", "aslB", "rpoB", "rpoC", "yjjZ", "rsmC",
#                       "yjtD", "thrA")
# 
# all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))
# 
# # --- Desired sample order ---
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )
# 
# # --- Filter, simplify, and mark presence ---
# blastedvariants_simple <- blastedvariants2 %>%
#   select(Gene.Symbol, Sample, qseqid) %>%
#   filter(Gene.Symbol %in% all_genes) %>%
#   distinct() %>%
#   mutate(
#     Category = case_when(
#       Gene.Symbol %in% biofilm_genes ~ "Biofilm",
#       Gene.Symbol %in% efflux_genes ~ "Efflux or Transport",
#       Gene.Symbol %in% persistent_genes ~ "Persistent shared",
#       TRUE ~ NA_character_
#     ),
#     Evolution_conditions = case_when(
#       grepl("^AZ", Sample) ~ "Azithromycin",
#       grepl("^CN", Sample) ~ "No Azithromycin",
#       TRUE ~ NA_character_
#     ),
#     Present = !is.na(qseqid)
#   ) %>%
#   complete(
#     Gene.Symbol = all_genes,
#     Sample = desired_samples,
#     fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
#   ) %>%
#   mutate(Sample = factor(Sample, levels = desired_samples))
# 
# # --- Duplicate fadL into both categories ---
# fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
# fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux or Transport")
# fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared",
#                                         Gene.Symbol = paste0(Gene.Symbol, "\u200B"))
# 
# blastedvariants_simple <- blastedvariants_simple %>%
#   filter(Gene.Symbol != "fadL") %>%
#   bind_rows(fadL_efflux, fadL_persistent)
# 
# # --- Ordering and facet label adjustment ---
# blastedvariants_simple <- blastedvariants_simple %>%
#   mutate(
#     Gene.Symbol = case_when(
#       Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
#       Category == "Efflux or Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
#       Category == "Persistent shared" ~ factor(Gene.Symbol,
#                                                levels = c(persistent_genes[persistent_genes != "fadL"], "fadL\u200B")),
#       TRUE ~ factor(Gene.Symbol)
#     ),
#     Category = case_when(
#       Category == "Efflux or Transport" ~ "Efflux or\nTransport",
#       TRUE ~ Category
#     )
#   )
# 
# # --- Plot with dark facet borders and updated legend ---
# ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
#   geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
#             color = "gray50", size = 0.5) +
#   scale_fill_manual(
#     values = c("Azithromycin" = "#FFC107", "No Azithromycin" = "#1E88E5"),
#     na.value = "white",
#     labels = c("Azithromycin", "No Azithromycin")  # explicitly rename legend entries
#   ) +
#   labs(
#     title = "Genes with AZ or CN variants by category",
#     x = "Gene",
#     y = "Sample",
#     fill = "Evolution condition"
#   ) +
#   facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
#   theme_minimal() +
#   theme(
#     panel.background = element_rect(fill = "white", color = "black"),
#     panel.border = element_rect(color = "black", fill = NA, size = 1),
#     plot.background = element_rect(fill = "white", color = NA),
#     panel.grid.major = element_line(color = "gray80"),
#     panel.grid.minor = element_line(color = "gray90"),
#     axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
#     axis.text.y = element_text(size = 10),
#     strip.text.x = element_text(size = 9, face = "bold"),
#     strip.background = element_rect(fill = "white", color = "black", size = 0.5)
#   )
# 
# 
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Genes of interest by category ---
biofilm_genes <- c("ycgV", "csgA", "csgC", "fimC", "fimF", "fimI",
                   "ybjK", "wcaC", "wcaD", "fliS", "fliT", "ampH")
efflux_genes <- c("sbmA", "acrR", "kefA", "yifK", "yjhB", "fadL")
persistent_genes <- c("frmA", "frmR", "yaiO", "yfcA", "mepA", "stfP", "tfaP", "tfaE",
                      "stfE", "pinE", "aslB", "yjtD", "rpoB", "rpoC", "rsmC",
                      "fadL", "thrA", "yfdF", "yjjZ")

all_genes <- unique(c(biofilm_genes, efflux_genes, persistent_genes))

# --- Desired sample order ---
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# --- Filter, simplify, and mark presence ---
blastedvariants_simple <- blastedvariants2 %>%
  select(Gene.Symbol, Sample, qseqid) %>%
  filter(Gene.Symbol %in% all_genes) %>%
  distinct() %>%
  mutate(
    Category = case_when(
      Gene.Symbol %in% biofilm_genes ~ "Biofilm",
      Gene.Symbol %in% efflux_genes ~ "Efflux or Transport",
      Gene.Symbol %in% persistent_genes ~ "Persistent shared",
      TRUE ~ NA_character_
    ),
    Evolution_conditions = case_when(
      grepl("^AZ", Sample) ~ "Azithromycin",
      grepl("^CN", Sample) ~ "No Azithromycin",
      TRUE ~ NA_character_
    ),
    Present = !is.na(qseqid)
  ) %>%
  complete(
    Gene.Symbol = all_genes,
    Sample = desired_samples,
    fill = list(Evolution_conditions = NA, Present = FALSE, Category = NA)
  ) %>%
  mutate(Sample = factor(Sample, levels = desired_samples))

# --- Duplicate fadL into both categories ---
fadL_rows <- blastedvariants_simple %>% filter(Gene.Symbol == "fadL")
fadL_efflux <- fadL_rows %>% mutate(Category = "Efflux or Transport")
fadL_persistent <- fadL_rows %>% mutate(Category = "Persistent shared",
                                        Gene.Symbol = "fadL\u200B")  # zero-width space

blastedvariants_simple <- blastedvariants_simple %>%
  filter(Gene.Symbol != "fadL") %>%
  bind_rows(fadL_efflux, fadL_persistent)

# --- Ordering and facet label adjustment ---
persistent_genes_ordered <- c("frmA", "frmR", "yaiO", "yfcA", "mepA", "stfP", "tfaP", "tfaE",
                              "stfE", "pinE", "aslB", "yjtD", "rpoB", "rpoC", "rsmC",
                              "fadL\u200B", "thrA", "yfdF", "yjjZ")

blastedvariants_simple <- blastedvariants_simple %>%
  mutate(
    Gene.Symbol = case_when(
      Category == "Biofilm" ~ factor(Gene.Symbol, levels = biofilm_genes),
      Category == "Efflux or Transport" ~ factor(Gene.Symbol, levels = efflux_genes),
      Category == "Persistent shared" ~ factor(Gene.Symbol, levels = persistent_genes_ordered),
      TRUE ~ factor(Gene.Symbol)
    ),
    Category = case_when(
      Category == "Efflux or Transport" ~ "Efflux or\nTransport",
      TRUE ~ Category
    )
  )

# --- Plot with dark facet borders and updated legend ---
plotfinal <- ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
            color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "No Azithromycin" = "#1E88E5"),
    na.value = "white",
    labels = c("Azithromycin", "No Azithromycin")  # legend labels
  ) +
  labs(
    title = "Genes with AZ or CN variants by category",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 9, face = "bold"),
    strip.background = element_rect(fill = "white", color = "black", size = 0.5)
  )
# --- Save the plot as a high-resolution PNG ---
plotfinal
# ggsave(
#   filename = "genes_variant_heatmap.png",   # file name
#   plot = last_plot(),                       # saves the last plot
#   width = 12,                               # width in inches
#   height = 6,                               # height in inches
#   dpi = 300  )
```

```{r}

plotfinal <- ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
            color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "No Azithromycin" = "#1E88E5"),
    na.value = "white",
    labels = c("Azithromycin", "No Azithromycin")
  ) +
  labs(
    title = "Genes with AZ or CN variants by category",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(
      angle = 90, hjust = 1, vjust = 0.5, size = 7,
      face = "italic"   # <-- make gene names italic
    ),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 9, face = "bold"),
    strip.background = element_rect(fill = "white", color = "black", size = 0.5)
  )
plotfinal
# ggsave(
#   filename = "genes_variant_heatmap.png",   # file name
#   plot = last_plot(),                       # saves the last plot
#   width = 12,                               # width in inches
#   height = 6,                               # height in inches
#   dpi = 300  )

```

```{r}

plotfinal <- ggplot(blastedvariants_simple, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = ifelse(Present, Evolution_conditions, NA)),
            color = "gray50", size = 0.5) +
  scale_fill_manual(
    values = c("Azithromycin" = "#FFC107", "No Azithromycin" = "#1E88E5"),
    na.value = "white",
    labels = c("Azithromycin", "No Azithromycin")
  ) +
  labs(
    title = "Genes with AZ or CN variants by category",
    x = "Gene",
    y = "Sample",
    fill = "Evolution condition"
  ) +
  facet_grid(. ~ Category, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_line(color = "gray90"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7, face = "italic"),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 9, face = "bold"),
    strip.background = element_rect(fill = "white", color = "black", size = 0.5)
  ) +
  # Add horizontal lines every 3 units
  geom_hline(yintercept = seq(3.5, length(unique(blastedvariants_simple$Sample)), by = 3), 
             color = "black", linetype = "solid", size = 1)


plotfinal

ggsave(
  filename = "genes_variant_heatmap.png",   # file name
  plot = last_plot(),                       # saves the last plot
  width = 12,                               # width in inches
  height = 6,                               # height in inches
  dpi = 300  )

```

