---
title: "Coding_ranges"
author: "Grace Carey"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(tidyverse)
library(dplyr)
library(tidyr)
```

First read in your annotated reference genome and the variants you are interested in. 
Make sure your files are quality filtered and ready to go. 

```{r, get files ready and quality filter}

annotat <- read.csv("MG1655_flye_annotation.csv") #this file contains the annotation for the MG1655 reference strain assembled/annotated genome
variants_df1 <- read.csv("combined_variations.csv")

# head(annotat)

variants_df <- dplyr::filter(variants_df1, Score >= 20)

cds_data <- annotat[annotat$type == "CDS", ]
names(cds_data)[names(cds_data) == "function."] <- "region_function"

```

Then use GenomicRanges to find where coding ranges (CDS) are found in the reference genome. 

```{r, find CDS ranges}
# Ensure start is always less than or equal to stop
cds_data$start_fixed <- pmin(cds_data$start, cds_data$stop)
cds_data$stop_fixed <- pmax(cds_data$start, cds_data$stop)


library(GenomicRanges)

cds_gr <- GRanges(
  seqnames = cds_data$contig_id,
  ranges = IRanges(start = cds_data$start_fixed, end = cds_data$stop_fixed),
  strand = cds_data$strand
)

mcols(cds_gr)$feature_id <- cds_data$feature_id
mcols(cds_gr)$region_function <- cds_data$region_function

```

Next find which variants of interest exist within a coding region (CDS). 

```{r, find which variants are in cds}

library(GenomicRanges)

variants_gr <- GRanges(
  seqnames = variants_df$Contig,
  ranges = IRanges(start = variants_df$Pos, width = 1)
)
# unique(seqnames(variants_gr)) #use this to check what the contig name is for the variants data

# Rename contig name in cds_gr to match the variant contig name
seqlevels(cds_gr) <- "562.163959.con.0001"

hits <- findOverlaps(variants_gr, cds_gr) #which variants are in a CDS?

# Extract overlapping variants from variants_df
variants_in_cds <- variants_df[queryHits(hits), ] #get the variants in CDS

# Extract CDS metadata (e.g., feature_id, region_function)
cds_hits <- cds_gr[subjectHits(hits)]
cds_meta <- as.data.frame(mcols(cds_hits))

# Combine both
variants_in_cds$feature_id <- cds_meta$feature_id
variants_in_cds$region_function <- cds_meta$region_function #now you have the desired variants


# write.csv(variants_in_cds, "variants_in_CDS.csv", row.names = FALSE) #can write to csv if needed. 

```

And finally find which variants exist within 500 basepairs (upstream or downstream) of a coding region. 

```{r, variants within 500bp of CDS}

# Expand CDS ranges by 500bp on either side
cds_gr_expanded <- resize(cds_gr, width = width(cds_gr) + 1000, fix = "center")

# Find overlaps between variants and the expanded CDS ranges
hits_expanded <- findOverlaps(variants_gr, cds_gr_expanded)

# Extract overlapping variants
variants_near_cds <- variants_df[queryHits(hits_expanded), ] # get the variants near CDS

# Extract expanded CDS metadata
cds_hits_expanded <- cds_gr_expanded[subjectHits(hits_expanded)]
cds_meta_expanded <- as.data.frame(mcols(cds_hits_expanded))

# Combine both
variants_near_cds$feature_id <- cds_meta_expanded$feature_id
variants_near_cds$region_function <- cds_meta_expanded$region_function # variants near CDS regions

# Optionally, save to a CSV
# write.csv(variants_near_cds, "variants_near_CDS.csv", row.names = FALSE)

# Split Var column by comma and expand the rows accordingly
variants_df_split <- variants_near_cds %>%
  separate_rows(Var, sep = ",")  # This will separate Var into individual rows

# Check the result
print(variants_df_split)
```

Now that we have the variants within 500bp of a coding region, I will see which (non-synonymous) ones are unique to AZ cultures, which are unique to control cultures, and which are common to both. Unique will be defined as:
Unique in position AND sequence. Since multiple variants can occur at a single position. 


```{r}
library(dplyr)
library(writexl)

# Read the key file
Key <- read.csv("SNPs_key.csv")

# Merge with variants data
merged_variants <- merge(variants_df_split, Key, by = "Library_name", all.x = TRUE)

# Create Pos_var to uniquely identify variants
merged_variants <- merged_variants %>%
  mutate(Pos_var = paste(Pos, Var, sep = "_"))

merged_variantsnonsyn <- filter(merged_variants, Type != "Synon")


```

```{r}

#I am commenting this out so that it doesn't keep adding pipes to the feature_id column. This code does work, just uncomment when ready to use. 
# blastresults <- read.csv("filteredblastinfo52125.csv") #these blast results are the result of blasting the unique variants against the nucleotide database to try and find more about what they actually are. After blasting they were quality filtered. These are the quality filtered blast results. 
# 
# merged_variantsnonsyn$feature_id <- paste0(merged_variantsnonsyn$feature_id, "|")
# 
# blastedvariants <- merge(blastresults, merged_variantsnonsyn,
#                          by.x = "qseqid", by.y = "feature_id", all = TRUE)

# write.csv(blastedvariants, "variantswithblastfiltered.csv")

```


```{r}

blastedvariants <- read.csv("variantswithblastfiltered.csv")

#in this dataframe, my own personal variant ids are in qseqid, and the feature IDs that BV BRC identified are in result.feature.id

length(unique(blastedvariants$feature_id))
length(unique(blastedvariants$qseqid))
sum(is.na(blastedvariants$subclass))


```

```{r}




library(ggplot2)
library(dplyr)
# Define the custom order of samples
custom_order <- c("CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
"AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
"AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
"AZ01_S07", "AZ01_S04", "AZ01_S01")

library(dplyr)
library(ggplot2)
library(scales)

# Step 1: Create P1 bins
blastedvariants <- blastedvariants %>%
  mutate(P1_bin = floor(Pos / 100000) * 100000)

# Step 2: Count occurrences by sample and P1_bin
heatmap_data <- blastedvariants %>%
  group_by(Culture, Timepoint, Sample, P1_bin) %>%
  summarise(count = n(), .groups = "drop")

# Step 3: Keep sample factor order
heatmap_data$Sample <- factor(heatmap_data$Sample, levels = custom_order)

# Plot using summarized dataframe
ggplot(heatmap_data, aes(x = P1_bin, y = Sample, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  scale_x_continuous(
    breaks = seq(0, 5000000, by = 1000000),
    labels = scientific,
    expand = c(0, 0)
  ) +
  labs(
    x = expression("Count of variations at reference genome positions (binned by 10" ^ 5 * " bp)"),
    y = "Sample",
    fill = "Count",
    title = expression("Count of variations at reference genome positions (binned by 10" ^ 5 * " bp)")
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  geom_vline(xintercept = seq(50000, 4650000, by = 100000), color = "gray", linetype = "solid")

# ggsave("variant_binned_map.png", width = 8, height = 6, dpi = 300)


```


```{r}

custom_order <- c("CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
"AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
"AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
"AZ01_S07", "AZ01_S04", "AZ01_S01")


blastedvariants$Sample <- factor(blastedvariants$Sample, levels = custom_order)
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_point(aes(color = Type), size = 2, alpha = 0.7) +
    scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Variants by position and category",
    x = "Reference Base Number",
    y = "Sample",
    color = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3))


```

```{r}
custom_order <- c("AZ01_S01", "AZ01_S04", "AZ01_S07",
                  "AZ02_S01", "AZ02_S04", "AZ02_S07",
                  "AZ03_S01", "AZ03_S04", "AZ03_S07",
                  "AZ04_S01", "AZ04_S04", "AZ04_S07",
                  "AZ05_S01", "AZ05_S04", "AZ05_S07",
                  "CN01_S01", "CN01_S04", "CN01_S07",
                  "CN02_S01", "CN02_S04", "CN02_S07")



blastedvariants$Sample <- factor(blastedvariants$Sample, levels = custom_order)
ggplot(blastedvariants, aes(x = Sample, fill = Type)) +
  geom_bar(position = "stack") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Count of Variant Types per Sample",
    x = "Sample",
    y = "Count",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9)) +
    geom_vline(xintercept = c(0.5,3.5,6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid")

# ggsave("variant_counts_bars.png", width = 8, height = 6, dpi = 300)


```

```{r}
library(dplyr)
library(ggplot2)

# Count per Timepoint, Type, and Culture
count_data <- blastedvariants %>%
  group_by(Timepoint, Type, Culture) %>%
  summarise(Count = n(), .groups = "drop")

# Plot: Faceted by Type, lines connecting points by Culture and Type, point color by Culture
ggplot(count_data, aes(x = Timepoint, y = Count, color = Culture, group = interaction(Culture, Type))) +
  geom_point(size = 3) +
  geom_line(aes(group = interaction(Culture, Type)), alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Variant Counts by Culture, Type, and Timepoint",
    x = "Timepoint",
    y = "Count",
    color = "Culture"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  ) +
  facet_wrap(~ Type) +  # Facet by Type
  geom_vline(xintercept = unique(count_data$Timepoint), color = "black", linetype = "solid")


```

```{r}
library(dplyr)
library(ggplot2)

# Count per Timepoint, Type, and Culture
count_data <- blastedvariants %>%
  group_by(Timepoint, Type, Evolution_conditions) %>%
  summarise(Count = n(), .groups = "drop")

# Add a new data point (Timepoint = S01, Type = Synon, Count = 0)
new_point <- data.frame(
  Timepoint = "S01", 
  Type = "Synon", 
  Evolution_conditions = "Azithromycin", # Replace with the actual condition if necessary
  Count = 0
)

# Add a new data point (Timepoint = S01, Type = Synon, Count = 0)
new_point1 <- data.frame(
  Timepoint = "S01", 
  Type = "Synon", 
  Evolution_conditions = "NoAzithromycin", # Replace with the actual condition if necessary
  Count = 0
)
# Add a new data point (Timepoint = S01, Type = Synon, Count = 0)
new_point2 <- data.frame(
  Timepoint = "S04", 
  Type = "Synon", 
  Evolution_conditions = "NoAzithromycin", # Replace with the actual condition if necessary
  Count = 0
)
# Add a new data point (Timepoint = S01, Type = Synon, Count = 0)
new_point3 <- data.frame(
  Timepoint = "S07", 
  Type = "Synon", 
  Evolution_conditions = "NoAzithromycin", # Replace with the actual condition if necessary
  Count = 0
)

# Combine the new point with the existing count_data
count_data <- rbind(count_data, new_point, new_point1, new_point2, new_point3)

# Plot: Faceted by Evolution_conditions, lines connecting points by Evolution_conditions and Type, point color by Type
ggplot(count_data, aes(x = Timepoint, y = Count, shape = Evolution_conditions, color = Type, group = interaction(Evolution_conditions, Type))) +
  geom_point(size = 3) +
  geom_line(aes(group = interaction(Evolution_conditions, Type)), alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  facet_grid(. ~ Evolution_conditions) +
  labs(
    title = "Variant Counts by Culture, Type, and Timepoint",
    x = "Timepoint",
    y = "Count",
    color = "Type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  )

```

```{r}
library(ggplot2)

# Custom order for Sample factor levels
custom_order <- c("CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
                  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
                  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
                  "AZ01_S07", "AZ01_S04", "AZ01_S01")

# Ensure Sample is a factor with the custom order
blastedvariants$Sample <- factor(blastedvariants$Sample, levels = custom_order)

# Plot with color gradient for Var_frac and shape for Type
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_point(aes(color = Var_frac, shape = Type), size = 2, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient from blue to red
  scale_shape_manual(values = 1:10) +  # Optional: Adjust number of shapes for each Type
  labs(
    title = "Variants by Position and Category",
    x = "Reference Base Number",
    y = "Sample",
    color = "Variant Fraction",
    shape = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3))

```

```{r}
# Plot with color gradient for Var_frac and shape for Type
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_point(aes(color = Var_frac), size = 2, alpha = 0.7) +  # Use Var_frac for color
  scale_color_gradient(low = "blue", high = "red") +  # Color gradient from blue to red
  labs(
    title = "Variants by Position and Sample",
    x = expression("Reference Genome Position (binned by 10" ^ 5 * " bp)"),
    y = "Sample",
    color = "Variant Fraction"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),  # Rotate x-axis labels
    axis.text.y = element_text(size = 8)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))  # Display only bins with data


```

```{r}

# Plot with color gradient for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_tile(aes(fill = Var_frac), color = "white") +  # Use Var_frac for fill in a heatmap
  scale_fill_gradient(low = "blue", high = "red") +  # Color gradient from blue to red
  labs(
    title = "Variant Fractions (Heatmap)",
    x = "Reference Base Number",
    y = "Sample",
    fill = "Variant Fraction"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))  # Display only positions with data

# ggsave("variant_frac_map.png", width = 8, height = 6, dpi = 300)


```

```{r}

# Plot with categorical color palette for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_tile(aes(fill = Type), color = "white") +  # Use Type for fill in a heatmap
  scale_fill_brewer(palette = "Dark2") +  # Apply categorical color palette for Type
    # scale_fill_brewer(palette = "Accent") +  # Apply categorical color palette for Type
  labs(
    title = "Variants by Position and Type",
    x = "Reference Base Number",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))  # Display only positions with data

# ggsave("variant_types_map.png", width = 8, height = 6, dpi = 300)

```


```{r}

# Plot with color gradient for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_tile(aes(fill = Mean_base_coverage), color = "white") +  # Use Var_frac for fill in a heatmap
  scale_fill_gradient(low = "blue", high = "red") +  # Color gradient from blue to red
  labs(
    title = "Coverage at Variant Positions (Heatmap)",
    x = "Reference Base Number",
    y = "Sample",
    fill = "Coverage"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))  # Display only positions with data

# ggsave("variant_coverage_map.png", width = 8, height = 6, dpi = 300)




```

```{r}

# Plot with color gradient for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  geom_tile(aes(fill = Score), color = "white") +  # Use Var_frac for fill in a heatmap
  scale_fill_gradient(low = "blue", high = "red") +  # Color gradient from blue to red
  labs(
    title = "Variant quality score (Heatmap)",
    x = "Reference Base Number",
    y = "Sample",
    fill = "Quality score"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))  # Display only positions with data

# ggsave("variant_frac_map.png", width = 8, height = 6, dpi = 300)




```
Now let's see where the Rhsa and b are: 



```{r}
rhs_subset <- datakeyfiltered[blastedvariants$region_function %in% c("Protein RhsA", "Protein RhsB"), ]

library(ggplot2)

# Add a logical column for highlighting
blastedvariants$highlight_rhsA <- blastedvariants$Function == "Protein RhsA"

library(ggplot2)

# Create a new column for highlighting category
blastedvariants$highlight_group <- ifelse(blastedvariants$Function == "Protein RhsA", "RhsA",
                                   ifelse(blastedvariants$Function == "Protein RhsB", "RhsB", NA))

ggplot(blastedvariants, aes(x = factor(Pos), y = Sample)) +
  # Base heatmap fill by variant Type
  geom_tile(aes(fill = Type), color = "white") +

  # Red border for RhsA
  geom_tile(
    data = subset(blastedvariants, highlight_group == "RhsA"),
    aes(x = factor(Pos), y = Sample),
    color = "hotpink", size = 2, fill = NA
  ) +

  # Blue border for RhsB
  geom_tile(
    data = subset(blastedvariants, highlight_group == "RhsB"),
    aes(x = factor(Pos), y = Sample),
    color = "yellow", size = 2, fill = NA
  ) +

  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Variants by Position and Type",
    x = "Reference Base Number",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") +
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 3),
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_discrete(breaks = unique(blastedvariants$Pos))


# ggsave("variant_highlightrhsa_map.png", width = 8, height = 6, dpi = 300)

```



```{r}

library(dplyr)
sum(is.na(blastedvariants$Gene.Symbol) | blastedvariants$Gene.Symbol == "")

# Step 1: Create a table of unique Gene.Symbol and Evolution_conditions pairs
gene_conditions <- blastedvariants %>%
  select(Gene.Symbol, Evolution_conditions) %>%
  distinct()

# Step 2: Create vectors for each condition
genes_azithro <- gene_conditions %>%
  filter(Evolution_conditions == "Azithromycin") %>%
  pull(Gene.Symbol)

genes_noazithro <- gene_conditions %>%
  filter(Evolution_conditions == "NoAzithromycin") %>%
  pull(Gene.Symbol)

# Step 3: Compare
only_azithro <- setdiff(genes_azithro, genes_noazithro)
only_noazithro <- setdiff(genes_noazithro, genes_azithro)
shared_genes <- intersect(genes_azithro, genes_noazithro)

# Step 4: Optionally, wrap in a list for clarity
gene_lists <- list(
  Only_in_Azithromycin = only_azithro,
  Only_in_NoAzithromycin = only_noazithro,
  Present_in_Both = shared_genes
)

# Print result
gene_lists

only_azithro_df <- as.data.frame(only_azithro)
only_noazithro_df <- as.data.frame(only_noazithro)
shared_genes_df <- as.data.frame(shared_genes)

library(openxlsx)

# Create a new workbook
wb <- createWorkbook()

# Add sheets and write data frames
addWorksheet(wb, "Only_Azithro")
writeData(wb, "Only_Azithro", only_azithro_df)

addWorksheet(wb, "Only_NoAzithro")
writeData(wb, "Only_NoAzithro", only_noazithro_df)

addWorksheet(wb, "Shared_Genes")
writeData(wb, "Shared_Genes", shared_genes_df)

# Save workbook to file
saveWorkbook(wb, file = "genes_location_summary.xlsx", overwrite = TRUE)


```

```{r}
library(dplyr)

blastedvariants <- blastedvariants %>%
  filter(!is.na(Gene.Symbol) & Gene.Symbol != "")


# Variants only found in NoAzithromycin
Variants_Control <- blastedvariants %>%
  filter(Gene.Symbol %in% only_noazithro & Evolution_conditions == "NoAzithromycin")

# Variants only found in Azithromycin
Variants_Azithro <- blastedvariants %>%
  filter(Gene.Symbol %in% only_azithro & Evolution_conditions == "Azithromycin")

# Variants shared between both conditions
Variants_Overlap <- blastedvariants %>%
  filter(Gene.Symbol %in% shared_genes & Evolution_conditions %in% c("Azithromycin", "NoAzithromycin"))


```

```{r}

# Plot with categorical color palette for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "white") +  # Use Type for fill in a heatmap
  scale_fill_brewer(palette = "Dark2") +  # Apply categorical color palette for Type
    # scale_fill_brewer(palette = "Accent") +  # Apply categorical color palette for Type
  labs(
    title = "Genes with variants",
    x = "Gene",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) 

# ggsave("variant_types_blasted_map.png", width = 10, height = 6, dpi = 300)

```


```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# # Desired Y-axis categories
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )

desired_samples <- c("AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# Ensure Gene.Symbol is a factor (or character)
Variants_Azithro$Gene.Symbol <- as.character(Variants_Azithro$Gene.Symbol)

# Expand the dataframe to include all combinations of Gene.Symbol and desired_samples
Variants_Azithro_complete <- Variants_Azithro %>%
  complete(Sample = desired_samples, Gene.Symbol, fill = list(Type = NA))

# Set Sample as an ordered factor for y-axis
Variants_Azithro_complete$Sample <- factor(Variants_Azithro_complete$Sample, levels = desired_samples)

# Plot
ggplot(Variants_Azithro_complete, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "white", na.rm = FALSE) +
scale_fill_brewer(palette = "Dark2", na.value = "#00000000", na.translate = FALSE) +
  labs(
    title = "Genes with variants: unique to azithromycin samples",
    x = "Gene",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) 

# ggsave("variant_types_blastedAZ_map.png", width = 10, height = 6, dpi = 300)

```


```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# # Desired Y-axis categories
# desired_samples <- c(
#   "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
#   "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
#   "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
#   "AZ01_S07", "AZ01_S04", "AZ01_S01"
# )

desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01")

# Ensure Gene.Symbol is a factor (or character)
Variants_Control$Gene.Symbol <- as.character(Variants_Control$Gene.Symbol)

# Expand the dataframe to include all combinations of Gene.Symbol and desired_samples
Variants_Control_complete <- Variants_Control %>%
  complete(Sample = desired_samples, Gene.Symbol, fill = list(Type = NA))

# Set Sample as an ordered factor for y-axis
Variants_Control_complete$Sample <- factor(Variants_Control_complete$Sample, levels = desired_samples)

# Plot
ggplot(Variants_Control_complete, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "white", na.rm = FALSE) +
scale_fill_brewer(palette = "Dark2", na.value = "#00000000", na.translate = FALSE) +
  labs(
    title = "Genes with variants: unique to control samples",
    x = "Gene",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) 

# ggsave("variant_types_blastedCN_map.png", width = 10, height = 4, dpi = 300)

```

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# # Desired Y-axis categories
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)


# Ensure Gene.Symbol is a factor (or character)
Variants_Overlap$Gene.Symbol <- as.character(Variants_Overlap$Gene.Symbol)

# Expand the dataframe to include all combinations of Gene.Symbol and desired_samples
Variants_Overlap_complete <- Variants_Overlap %>%
  complete(Sample = desired_samples, Gene.Symbol, fill = list(Type = NA))

# Set Sample as an ordered factor for y-axis
Variants_Overlap_complete$Sample <- factor(Variants_Overlap_complete$Sample, levels = desired_samples)

# Plot
ggplot(Variants_Overlap_complete, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = Type), color = "white", na.rm = FALSE) +
scale_fill_brewer(palette = "Dark2", na.value = "#00000000", na.translate = FALSE) +
  labs(
    title = "Genes with variants: present in both AZ and control samples",
    x = "Gene",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) 

# ggsave("variant_types_blastedOVERLAP_map.png", width = 10, height = 6, dpi = 300)

```


Now let's try to look at systems and subsystems

```{r}

c25 <- c(
  "dodgerblue2", "#E31A1C", # red
  "yellow4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "#FDBF6F", "red",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "brown", # lt purple
  "gray70", # lt orange
  "black", "orange",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "green4", "yellow3",
  "darkorange4", "gray90"
)
# pie(rep(1, 25), col = c25)
blastedvariants$subclass[blastedvariants$subclass == ""] <- NA

# Plot with categorical color palette for Var_frac as a heatmap
ggplot(blastedvariants, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = subclass), color = "white") +  # Use Type for fill in a heatmap
  scale_fill_manual(values = c25) +  # Use the custom palette here
    # scale_fill_brewer(palette = "Accent") +  # Apply categorical color palette for Type
  labs(
    title = "Genes with variants",
    x = "Gene",
    y = "Sample",
    fill = "Variation Type"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) 

# ggsave("variant_subclass_blasted_map.png", width = 14, height = 6, dpi = 300)

```

```{r}
library(dplyr)
library(ggplot2)

# Your desired samples and their order
# desired_samples <- c("CN01_S01", "CN01_S04", "CN01_S07", "CN02_S01", "CN02_S04", "CN02_S07")
desired_samples <- c(   "CN02_S07", "CN02_S04", "CN02_S01","CN01_S07","CN01_S04","CN01_S01")

# Clean subclass column
Variants_Control$subclass[Variants_Control$subclass == ""] <- NA
Variants_Control$Gene.Symbol <- as.character(Variants_Control$Gene.Symbol)

# Find missing samples
existing_samples <- unique(Variants_Control$Sample)
missing_samples <- setdiff(desired_samples, existing_samples)

# Create dummy rows for missing samples to show on y-axis, with a placeholder gene that won't show on x-axis
dummy_rows <- data.frame(
  Sample = missing_samples,
  Gene.Symbol = NA_character_,  # NA so no tile on x axis
  subclass = NA_character_,
  stringsAsFactors = FALSE
)

# Combine real and dummy rows
Variants_Control_filled <- bind_rows(Variants_Control, dummy_rows)

# Set factor levels for y axis (Samples)
Variants_Control_filled$Sample <- factor(Variants_Control_filled$Sample, levels = desired_samples)

# Set factor levels for x axis (Genes) if needed (optional)
# Variants_Control_filled$Gene.Symbol <- factor(Variants_Control_filled$Gene.Symbol)

ggplot(Variants_Control_filled, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = subclass), color = "white", na.rm = FALSE) +
  scale_fill_manual(
    values = c(
      "Flagellum" = "#1f77b4",
      "Arginine; urea cycle, creatine, polyamines" = "#ff7f0e",
      "RNA processing and modification" = "#2ca02c",
      "Aminoacyl-tRNA-synthetases" = "#d62728",
      "Pyrimidines" = "#9467bd"
    ),
    na.value = "gray80"  # color for missing subclass tiles
  ) +
  labs(
    title = "Genes with variants: unique to control samples",
    x = "Gene",
    y = "Sample",
    fill = "Subclass"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    axis.text.y = element_text(size = 10)
  )


# ggsave("variant_subclass_blastedCN_map.png", width = 10, height = 4, dpi = 300)

```



```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# # Desired Y-axis categories
desired_samples <- c(
  "CN02_S07", "CN02_S04", "CN02_S01", "CN01_S07", "CN01_S04", "CN01_S01",
  "AZ05_S07", "AZ05_S04", "AZ05_S01", "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01", "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)


# Ensure Gene.Symbol is a factor (or character)
Variants_Overlap$Gene.Symbol <- as.character(Variants_Overlap$Gene.Symbol)

Variants_Overlap$subclass[Variants_Overlap$subclass == ""] <- NA


ggplot(Variants_Overlap, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = subclass), color = "white", na.rm = FALSE) +
  scale_fill_manual(
    values = c25,
    na.value = "darkgray"  # Color for real NA subclass values
  ) +
  labs(
    title = "Genes with variants in both AZ and control samples",
    x = "Gene",
    y = "Sample",
    fill = "Subclass"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:25, color = "grey90", linetype = "solid") + 
  geom_hline(yintercept = c(3.5, 6.5, 9.5, 12.5, 15.5, 18.5, 21.5), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    axis.text.y = element_text(size = 10)
  )


# ggsave("variant_subclass_blastedOVERLAP_map.png", width = 10, height = 4, dpi = 300)

```

```{r}

library(dplyr)
library(ggplot2)

# Desired samples in the order you want on the y-axis
desired_samples <- c(
  "AZ05_S07", "AZ05_S04", "AZ05_S01",
  "AZ04_S07", "AZ04_S04", "AZ04_S01",
  "AZ03_S07", "AZ03_S04", "AZ03_S01",
  "AZ02_S07", "AZ02_S04", "AZ02_S01",
  "AZ01_S07", "AZ01_S04", "AZ01_S01"
)

# Clean subclass column
Variants_Azithro$subclass[Variants_Azithro$subclass == ""] <- NA
Variants_Azithro$Gene.Symbol <- as.character(Variants_Azithro$Gene.Symbol)

# Find missing samples
existing_samples <- unique(Variants_Azithro$Sample)
missing_samples <- setdiff(desired_samples, existing_samples)

# Create dummy rows for missing samples so they show on the y-axis
dummy_rows <- data.frame(
  Sample = missing_samples,
  Gene.Symbol = NA_character_,  # NA means no tile on x axis
  subclass = NA_character_,
  stringsAsFactors = FALSE
)

# Combine original and dummy rows
Variants_Azithro_filled <- bind_rows(Variants_Azithro, dummy_rows)

# Set factor levels for y-axis (Samples)
Variants_Azithro_filled$Sample <- factor(Variants_Azithro_filled$Sample, levels = desired_samples)

# Optional: set Gene.Symbol as factor if you want specific order on x-axis
# Variants_Azithro_filled$Gene.Symbol <- factor(Variants_Azithro_filled$Gene.Symbol)

# Define colors for subclasses - you can customize these hex codes
subclass_colors <- c(
  "Capsule and Slime layer" = "#1f77b4",
  "Multidrug efflux systems" = "#ff7f0e",
  "Branched-chain amino acids" = "#2ca02c",
  "Fermentation" = "#d62728",
  "Transcription" = "#9467bd",
  "Ribosome biogenesis" = "#8c564b",
  "RNA processing and modification" = "#e377c2",
  "Protein folding" = "#7f7f7f",
  "Aromatic amino acids and derivatives" = "#bcbd22"
)

ggplot(Variants_Azithro_filled, aes(x = Gene.Symbol, y = Sample)) +
  geom_tile(aes(fill = subclass), color = "white", na.rm = FALSE) +
  scale_fill_manual(
    values = subclass_colors,
    na.value = "gray80"  # color for missing subclass tiles
  ) +
  labs(
    title = "Genes with variants: unique to Azithromycin samples",
    x = "Gene",
    y = "Sample",
    fill = "Subclass"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 0.5:(length(desired_samples) + 5), color = "grey90", linetype = "solid") + 
  # Adjust or remove the next line if it doesn't fit your data grid
  geom_hline(yintercept = seq(3.5, length(desired_samples), by = 3), color = "black", linetype = "solid") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    axis.text.y = element_text(size = 10)
  )

ggsave("variant_subclass_blastedAZ_map.png", width = 10, height = 4, dpi = 300)

```


```{r}
# Overlap
# "yecD" "aspS"  "yneO" "ycjM" "pspE" "pspD" "pinE" "stfP" "tfaP" "tfaE" "stfE" "ymfQ" "ybaP" "ybaK" "yaiO" "frmR" "frmA" "ykgE" "ykgD"
# "thrA" "yjtD" "rsmC" "yjjZ" "yjhC" "yjhB" "rpoC" "rpoB" "aslB" "yifK" "rfaL" "rfaC" "ygaC" "ygaW" "yfdF" "fadL" "mepA" "yfcA"
# 
# Control only 
# "yedJ" "dcm"  "amyA" "fliT" "fliS" "ydhC" "ydhB" "ydgI" "ydgH" "rne"  "glnS" "nagE" "borD" "rzpD" "arrD" "yagF" "carB"
# 
# Azithromycin only 
# "wcaC" "wcaD" "ycgV" "csgC" "csgA" "ycbF" "fimF" "ybjK" "ybjJ" "kefA" "acrR" "sbmA" "ampH" "ilvI" "leuO" "fimC" "fimI" "rhsA" "pitA"
# "yhiN" "ykgK" "yqeG" "rpoA" "rplQ" "rluD" "yfiH" "clpB" "aroC" "yfbV" "yfbU"


genes_of_interestOVERLAP <- c(
  "yecD", "aspS", "yneO", "ycjM", "pspE", "pspD", "pinE", "stfP", "tfaP", "tfaE",
  "stfE", "ymfQ", "ybaP", "ybaK", "yaiO", "frmR", "frmA", "ykgE", "ykgD",
  "thrA", "yjtD", "rsmC", "yjjZ", "yjhC", "yjhB", "rpoC", "rpoB", "aslB",
  "yifK", "rfaL", "rfaC", "ygaC", "ygaW", "yfdF", "fadL", "mepA", "yfcA"
)

Overlap_products <- blastedvariants[blastedvariants$Gene.Symbol %in% genes_of_interestOVERLAP, c("Gene.Symbol", "Product")]

# View results
head(Overlap_products)


genes_of_interestCN <- c(
  "yedJ", "dcm",  "amyA", "fliT", "fliS", "ydhC", "ydhB", "ydgI", "ydgH", "rne",  "glnS", "nagE", "borD", "rzpD", "arrD", "yagF", "carB"
)

Control_products <- blastedvariants[blastedvariants$Gene.Symbol %in% genes_of_interestCN, c("Gene.Symbol", "Product")]

# View results
head(Control_products)


genes_of_interestAZ <- c(
  "wcaC", "wcaD", "ycgV", "csgC", "csgA", "ycbF", "fimF", "ybjK", "ybjJ", "kefA", "acrR", "sbmA", "ampH", "ilvI", "leuO", 
  "fimC", "fimI", "rhsA", "pitA", "yhiN", "ykgK", "yqeG", "rpoA", "rplQ", "rluD", "yfiH", "clpB", "aroC", "yfbV", "yfbU"
)


AZ_products <- blastedvariants[blastedvariants$Gene.Symbol %in% genes_of_interestAZ, c("Gene.Symbol", "Product")]

# View results
head(AZ_products)


library(dplyr)
library(openxlsx)

# Make distinct dataframes
Overlap_products_distinct <- distinct(Overlap_products)
Control_products_distinct <- distinct(Control_products)
AZ_products_distinct <- distinct(AZ_products)

# Create workbook
wb <- createWorkbook()

# Add sheets and write distinct data
addWorksheet(wb, "Overlap_products")
writeData(wb, "Overlap_products", Overlap_products_distinct)

addWorksheet(wb, "Control_products")
writeData(wb, "Control_products", Control_products_distinct)

addWorksheet(wb, "AZ_products")
writeData(wb, "AZ_products", AZ_products_distinct)

# Save workbook
saveWorkbook(wb, file = "products_combined_distinct.xlsx", overwrite = TRUE)


```
